---
title: "Lab 5 new"
author: "Julia Van Goor"
date: "10/07/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Owner/Documents/UMich_Bio201_F19/")
```

# Load packages

```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
set.seed(7)
```

# Relationship between data structure and results
There are several factors of data sets which influence the results of statistical tests: 

* Values: numeric values of each data point
* Sample size (n): number of data points in each group being compared 
* Variance: spread of the data within each group
* Effect size: size of the difference in mean, strength of the correlation, etc. 

Before getting into the details of use and interpretation of formal statistical tests, we are going to do an exercise with random number vectors to illustrate how each of these factors effects your intuitive interpretation of the results. 

Run the code blocks below to create two vectors with the listed means, sample size, variation (via SD). Combine these vectors into a data frame and plot. Discuss with your neighbor, then under each plot write if you would determine these two samples to be equal, and which factors influenced your decision. 

### Example 1
```{r}
sample1 <- rnorm(6, mean=12, sd = 1) #vector1 

sample2 <- rnorm(6, mean=15, sd = 1) #vector2

df1 <- cbind(sample1, sample2) %>% #combine vectors into dataframe
  as.data.frame() %>% 
  gather(key = "samples", value = "numbers") %>% #make long format
  mutate(samples = as.factor(samples)) #convert to factor for plot

ggplot(data = df1, aes(x = samples, #plot data frame 
                     y = numbers)) + 
  geom_boxplot(aes(color = samples)) + #shows median & quartiles 
  geom_jitter(aes(color = samples)) + #jitter = individual points per category 
  theme(legend.position = "none")
#I do not think that these samples are equal, because the one on the right has one less point than the one on the left, a much higher mean than the one on the left, and a much larger standard deviation than the one on the left
```

### Example 2
```{r}
sample3 <- rnorm(6, mean=12, sd = 3) 
sample4 <- rnorm(6, mean=15, sd = 4) 

df2 <- cbind(sample3, sample4) %>% #combine vectors into dataframe
  as.data.frame() %>% 
  gather(key = "samples", value = "numbers") %>% #make long format
  mutate(samples = as.factor(samples)) #convert to factor for plot

ggplot(data = df2, aes(x = samples, #plot data frame 
                     y = numbers)) + 
  geom_boxplot(aes(color = samples)) + 
  geom_jitter(aes(color = samples)) + 
  theme(legend.position = "none")
#I think that these samples are similar, but not exactly equal. They have the same number of data points and closer means and standard deviations, but sample 4 still has a higher mean than sample 3 and sample 3 apprears to have a slightly larger standard deviation.
```

### Example 3
```{r}
sample5 <- rnorm(20, mean=12, sd = 1) 
sample6 <- rnorm(20, mean=15, sd = 1) 

df3 <- cbind(sample5, sample6) %>% #combine vectors into dataframe
  as.data.frame() %>% 
  gather(key = "samples", value = "numbers") %>% #make long format
  mutate(samples = as.factor(samples)) #convert to factor for plot

ggplot(data = df3, aes(x = samples, #plot data frame 
                     y = numbers)) + 
  geom_boxplot(aes(color = samples)) + 
  geom_jitter(aes(color = samples)) + 
  theme(legend.position = "none")
#These samples are not the same, becuase they have different (but close) means, however they do have very similar standard deviations
```

### Example 4
```{r}
sample7 <- rnorm(20, mean=12, sd = 3) 
sample8 <- rnorm(20, mean=15, sd = 4) 

df4 <- cbind(sample7, sample8) %>% #combine vectors into dataframe
  as.data.frame() %>% 
  gather(key = "samples", value = "numbers") %>% #make long format
  mutate(samples = as.factor(samples)) #convert to factor for plot

ggplot(data = df4, aes(x = samples, #plot data frame 
                     y = numbers)) + 
  geom_boxplot(aes(color = samples)) + 
  geom_jitter(aes(color = samples)) + 
  theme(legend.position = "none")
#Again, these samples are similar but not "equal" becuase they have slightly different means even though the standard deviations look similar and overlap
```

# Import data 

Import the data frame you generated last week that has the mean weekly SCFAs. Import both long and wide formats of these data. During import convert column names to snake case.  
```{r include=FALSE}
scfa_long <- read_delim("Lab5/curated_data/SCFA_wkly_long.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
  

scfa_wide <- read_delim("Lab5/curated_data/SCFA_wkly_wide.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
```


# Introduction

As mentioned earlier this semester, one of the over arching questions for this course is: "Does the consumption of a prebiotic (fiber supplement) effect the gut microbiome?" As you have been learning there are many elements to this; fermentation products, community composition, pH, and host lifestyle. In most of the lab exercise today we will attempt to answer part of this question using the change in fermentation products (specifically butyrate) in individuals who consumed potato starch. Today we will introduce several types of plots, and discuss when they are appropriate for certain families of data, and we will conduct some basic statistical tests for these plots. 

```{r}
# use select + starts_with or ends_with to retain/drop multiple columns with similar names
butyrate_wide <- scfa_wide %>%
  select(-starts_with("ace"), -ends_with("ace"), 
         -starts_with("pro"), -ends_with("pro"),
         -starts_with("total"), -ends_with("total"),
         -starts_with("delta")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT") %>%
  na.omit(but_wk1, but_wk3)
```

# 1 continuous and 1 categorical (discrete) variable 

In this lesson we are going to use the term categorical, however these are called discrete on the ggplot cheatsheet. 

### Violin plots
In the section above we subset by fiber supplement and fermentation product. One of the variations of fiber supplement we tested was frequency (once or twice daily). In the code below we are going to generate a plot that illustrates the butyrate concentration when individuals are consuming fiber supplements at different frequencies. 

Violin plots (or geom_boxplot + geom_jitter) are the recommended way to represent these families of data, they show the variation and the range, are easy to annotate with a mean (geom_errobar), you can easily add individual data points (geom_violin + geom_jitter), and do not obfuscate data in the manner of a bar plot. We will not use bar plots in this course. 

Subset the long data frame imported above for butyrate measurements only, supplement type is potato starch (BRMPS or LOODAT), drop any samples from Winter 2015, filter for individuals who were quantity compliant, make sure to keep the frequency and semester columns. 
```{r}
butyrate_long <- scfa_long %>%
  select(-starts_with("ace"), -starts_with("pro"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(butyrate_mean) #drop any samples with NA in named columns

# Generate a plot of the [butyrate] of each supplement type, week 3 only  
butyrate_long %>%
  filter(study_week == "week3") %>%
  ggplot(aes(x = frequency, 
             y = butyrate_mean)) + 
  geom_violin() # + geom_jitter()
```

In the plots created above, notice the widest part of the twice daily category appears slightly higher than once daily. This means most of the data points in the 2x group are of a higher concentration that those in the 1x group. However the top of the violin in the 1x group is higher than the 2x, indicating the 1x group's maximum value is greater. Discuss with your neighbor, do you think the butyrate concentrations of these two groups (1x vs. 2x) are different? They appear to be similar, but it is possible that the 2x is slightly higher because the widest part is slightly higher than 1x, and 2x has a smaller range.

### Checking assumptions
In the sections above we observed trends in the data, but this is not sufficient for research purposes. Scientists use statistics to determine the probability that these trends are real. Before we can dive into using a statistical test, we have to determine if our data are appropriate for the test of interest, otherwise we might have more confidence in the results than we should. We do this by checking the assumptions of the tests. 

In the violin plot above, we want to determine if the butyrate concentrations in 1xdaily potato starch consumers is different from that of the 2xdaily potato starch consumers. The plot indicates this might be the case. To answer this question we are comparing two means, to do this we use the student's t-test. A t-test requires the following assumptions to be met: 

* Relatively large sample size (usually > 30)
* The samples have to come from a normal distribution
* We are also going to check if the variances of the groups are equal, because this will determine some of the t.test() arguments

##### Sample size
```{r}
# check sample size
butyrate_long %>%
  filter(study_week == "week3") %>% #subset long data frame in same manner as for plot
  group_by(frequency) %>% #set groups to what we are comparing
  summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group
```

##### Normality 
```{r}
# normal distribution 
but_df <- butyrate_long %>%
  filter(study_week == "week3") #subset long data frame in same manner as for plot

shapiro.test(but_df$butyrate_mean) #call column of df with values = vector 

```
Here a small p-value indicates these samples differ from a normal distribution. When I have a result with a samll p-value I always check how much the distrubtion deviates from normal with a histogram:

```{r}
ggplot(but_df, aes(x=butyrate_mean)) +
  geom_histogram() #histograms show the number of data points (count) at each value

qqnorm(but_df$butyrate_mean); qqline(but_df$butyrate_mean)
```
This histogram shows a rough bell curve, in combination with our large sample size we are okay with proceeding with a t-test. 

#### Equal variances 
Finally we are going to determine if the groups we would like to compare (1xdaily and 2xdaily) have equal variances (homogeneity). 
```{r}
# code to extract first group (1xdaily)
once_grp <- butyrate_long %>%
  filter(study_week == "week3",
         frequency == "1xdaily") 

# code to extract second group (2xdaily)
twice_grp <- butyrate_long %>%
  filter(study_week == "week3",
         frequency == "2xdaily") 
```

```{r}
var.test(x = once_grp$butyrate_mean, 
         y = twice_grp$butyrate_mean, 
         alternative = "two.sided")
```
A low p-value indicates the variances are not equal, we will account for this in our test in the next section. 

### T-test between categorical variables
The test you will probably use most frequently is the t-test; and this test determines if the means of two groups are equal. First we need to extract the data we will use for the test: 

```{r}
# use groups extracted above for the test 
t.test(x = once_grp$butyrate_mean, 
       y = twice_grp$butyrate_mean,
       alternative = "less", paired = FALSE, var.equal = FALSE)
```
A large p-value indicates the means of the two groups are not different. 

### Paired t-test
As we've discussed in Friday lectures, everyone has a unique microbiome. We should compare week 1 and week 3 concentrations of all individuals who consumed BRMPS, and determine if the addition of the supplement results in generally higher fecal butyrate concentrations for most individuals. This is accomplished with a special flavor of t-test called a paired t-test. Paired t-tests are used whenever the samples are not independent, such as when the samples are from the same individual over time. 

```{r}
# Same plot as above but use facets to plot both weeks 
butyrate_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = butyrate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + # geom_jitter() + 
  facet_grid(~frequency) + 
  xlab(NULL) + 
  ylab("Butyrate (mmol/kg)") + 
  theme(legend.position = "none")
```
In the figure above it appears there is no difference in weeks for the 1xdaily group, but there may be an increase in the 2xdaily group. 

Before we conduct the statistical test to determine if the observed trends are likely true, we must check our assumptions.
```{r}
# sample size
butyrate_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>%  #subset long data frame in same manner as plot
  group_by(frequency, study_week) %>% #set groups to what we are comparing
  summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group
```

```{r}
# Check assumptions for each week of the 2xdaily groups 
wk1_2x <- butyrate_long %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 
shapiro.test(wk1_2x$butyrate_mean) 
ggplot(wk1_2x, aes(x = butyrate_mean)) + geom_histogram()

wk3_2x <- butyrate_long %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "2xdaily") 
shapiro.test(wk3_2x$butyrate_mean) 
ggplot(wk3_2x, aes(x = butyrate_mean)) + geom_histogram()

# join these data frames back together
x2_df <- inner_join(x = wk1_2x, y = wk3_2x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(butyrate_mean_wk1 = butyrate_mean.x,
         butyrate_mean_wk3 = butyrate_mean.y) %>%
  select(-starts_with("study_week"))

# code to run paired t-test 
t.test(x = x2_df$butyrate_mean_wk1, y = x2_df$butyrate_mean_wk3, 
       alternative = "less", paired = TRUE) 
```
For this paired t-test we specified alternative = less because we expect the butyrate concentrations to be higher during week 3. Read the details of the t.test() help page for options for the alternative argument.

The p-value is 0.07 which for a complex biological system (like the gut microbiome) is low! The pattern we observed in the figure is likely a real trend. We can say with confidence that the butyrate concentrations between weeks 1 and 3 are not equal for individuals who consumed BRMPS twice daily. 

### Non-parametric test
What if our data set does not meet any of the assumptions for the test? We just use another test. A t-test is a parametric test, and the non-parametric counterpart is the Mann-Whitney-U test (also called a two-sample Wilcoxon test).
```{r}
# same arguments, just a different function call. 
wilcox.test(x = x2_df$butyrate_mean_wk1, 
            y = x2_df$butyrate_mean_wk3, 
            alternative = "less", paired = TRUE)
```

# Homework 5.1
Repeat the process to conduct the paired t-test for the 1xdaily group. Remember to check the assumptions. Write your conclusions regarding the test as a comment at the end of the code block. 
```{r}
# copy + paste code, and update column and dataframes names to run test
butyrate_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>%  #subset long data frame in same manner as plot
  group_by(frequency, study_week) %>% #set groups to what we are comparing
  summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group
wk1_1x <- butyrate_long %>%
  filter(study_week == "week1", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk1_1x$butyrate_mean) 
ggplot(wk1_1x, aes(x = butyrate_mean)) + geom_histogram()

wk3_1x <- butyrate_long %>%
  filter(study_week == "week3", 
         supplement_consumed == "BRMPS", 
         frequency == "1xdaily") 
shapiro.test(wk3_1x$butyrate_mean) 
ggplot(wk3_1x, aes(x = butyrate_mean)) + geom_histogram()

var.test(x = wk1_1x$butyrate_mean, 
         y = wk3_1x$butyrate_mean, 
         alternative = "two.sided")

# join these data frames back together
x1_df <- inner_join(x = wk1_1x, y = wk3_1x,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(butyrate_mean_wk1 = butyrate_mean.x,
         butyrate_mean_wk3 = butyrate_mean.y) %>%
  select(-starts_with("study_week"))

# code to run paired t-test 
t.test(x = x1_df$butyrate_mean_wk1, y = x1_df$butyrate_mean_wk3, 
       alternative = "less", paired = TRUE) 
#The 1x daily for week 1 and week 3 have high enough sample sizes and a roughly normal distributions even though they are skewed slightly to the right. Both of them have roughly equal variances, because the p value for the variance test is high (rougly 0.64). I believe it fits the assumptions for a t-test. The paired t-test has a high p-value of 0.2367 so we fail to reject the null hypothesis that that the butyrate concentration did not change significantly.
```

# Homework 5.2
Generate plots to determine if methane, acetate, propionate increased during week 3 of the study. Generate plot to determine if pH decreased during week 3 of the study. Save plots to folder called `figures`. 
```{r}
# breath methane
breath_wkly <- read_delim("curated_data/breath_wkly.txt", #import data
    "\t", escape_double = FALSE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
View(breath_wkly)
ch4_long <- breath_wkly %>% 
  select(-starts_with("h2"), -starts_with("co2"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(ch4_mean) #drop any samples with NA in named columns
View(ch4_long)
plot_ch4 <- ch4_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = ch4_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + # geom_jitter() + 
  facet_grid() + 
  xlab(NULL) + 
  ylab("Methane Concentration (ppm)") + 
  theme(legend.position = "none")
plot_ch4 #the range of methane measurements increased in week three however, it does not appear that the actual mean methane measurement increased from week 1

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_ch4_violin.pdf",
          plot = plot_ch4,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)
```

```{r}
# acetate
acetate_long <- scfa_long %>%
  select(-starts_with("but"), -starts_with("pro"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(acetate_mean) #drop any samples with NA in named columns
plot_ace <- acetate_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = acetate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + # geom_jitter() + 
  facet_grid() + 
  xlab(NULL) + 
  ylab("Acetate Concentration (mmol/kg)") + 
  theme(legend.position = "none")
plot_ace #it appears that maybe the concentration of acetate increased from week one to week three becuase the week three plot is wider at higher points

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_ace_violin.pdf",
          plot = plot_ace,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

```

```{r}
# propionate
propionate_long <- scfa_long %>%
  select(-starts_with("but"), -starts_with("ace"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(propionate_mean) #drop any samples with NA in named columns
plot_pro <- propionate_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = propionate_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + # geom_jitter() + 
  facet_grid() + 
  xlab(NULL) + 
  ylab("Propionate Concentration (mmol/kg)") + 
  theme(legend.position = "none")
plot_pro #it does not look like the amount of propionate increased from week one to week 3, because even though week 3 has a larger range than week 1, it has a larger number of measurements concentrated at a lower mean

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_pro_violin.pdf",
          plot = plot_pro,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)
```

```{r}
# pH
pH_wkly <- read_delim("curated_data/pH_wkly.txt", 
    "\t", escape_double = FALSE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
View(pH_wkly)
pH_long <- pH_wkly %>% 
  select(-ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(ph_mean) #drop any samples with NA in named columns
View(pH_long)
plot_pH <- pH_long %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = ph_mean, 
             color = study_week), 
         frequency) + 
  geom_violin() + # geom_jitter() + 
  facet_grid() + 
  xlab(NULL) + 
  ylab("Average pH") + 
  theme(legend.position = "none")
plot_pH #it does look possible that the pH could have decreased from week 1 to week 3, becuase the week 3 shows a larger range in the direction of lower numbers and more samples concentrated in these lower number regions.However, this shift is on an order of less than 1 pH point which may not be significant in the grand scheme of things.

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_pH_violin.pdf",
          plot = plot_pH,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)
```

# Homework 5.3
Check assumptions on data sets for individuals who consumed BRMPS: breath methane, pH, acetate, propionate
```{r}
# breath methane
group_by(ch4_long, study_week) %>% #set groups to what we are comparing
summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group

ch4_wk1 <- ch4_long %>%
  filter(study_week == "week1")
shapiro.test(ch4_wk1$ch4_mean) 
ggplot(ch4_wk1, aes(x = ch4_mean)) + geom_histogram()

ch4_wk3 <- ch4_long %>%
  filter(study_week == "week3")
shapiro.test(ch4_wk3$ch4_mean) 
ggplot(ch4_wk3, aes(x = ch4_mean)) + geom_histogram()

var.test(x = ch4_wk1$ch4_mean, 
         y = ch4_wk3$ch4_mean, 
         alternative = "two.sided")

#Both week 1 and week 3 for methane have well over 30 participants and are even in similar counts to eachother. Both of their distributions are unimodal and heavily skewed to the right. Additionally, they both fail the Shapiro-Wilk normality test with p values near 0. For this reason I chose to run the non-parametric test in addition to the paired T-test.
```

```{r}
# acetate
acetate_long <- scfa_long %>%
  select(-starts_with("but"), -starts_with("pro"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(acetate_mean) #drop any samples with NA in named columns

group_by(acetate_long, study_week) %>% #set groups to what we are comparing
summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group

ace_wk1 <- acetate_long %>%
  filter(study_week == "week1")
shapiro.test(ace_wk1$acetate_mean) 
ggplot(ace_wk1, aes(x = acetate_mean)) + geom_histogram()

ace_wk3 <- acetate_long %>%
  filter(study_week == "week3")
shapiro.test(ace_wk3$acetate_mean) 
ggplot(ace_wk3, aes(x = acetate_mean)) + geom_histogram()

var.test(x = ace_wk1$acetate_mean, 
         y = ace_wk3$acetate_mean, 
         alternative = "two.sided")
#The week1 and week3 data both appear to be unimodal and normal, even though they are skewed to the right. They also both have sample sizes well over 30 and similar variances. Since they appear approximately normal I will run a paired T-test.

```

```{r}
# propionate
propionate_long <- scfa_long %>%
  select(-starts_with("but"), -starts_with("ace"), -ends_with("median")) %>%
  filter(semester != "Winter2015",
         supplement_consumed == "BRMPS" | supplement_consumed == "LOODAT",
         quantity_compliant == "yes") %>%
  na.omit(propionate_mean) #drop any samples with NA in named columns

group_by(propionate_long, study_week) %>% #set groups to what we are comparing
summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group

pro_wk1 <- propionate_long %>%
  filter(study_week == "week1")
shapiro.test(pro_wk1$propionate_mean) 
ggplot(pro_wk1, aes(x = propionate_mean)) + geom_histogram()

pro_wk3 <- propionate_long %>%
  filter(study_week == "week3")
shapiro.test(pro_wk3$propionate_mean) 
ggplot(pro_wk3, aes(x = propionate_mean)) + geom_histogram()

var.test(x = pro_wk1$propionate_mean, 
         y = pro_wk3$propionate_mean, 
         alternative = "two.sided")
#The week1 and week3 data both appear to be normal, even though week 1 appears to have multiple modes, and week 3 is skewed to the right. Even though their variances are not very similar, they also both have sample sizes well over 30. Since they appear approximately normal, I believe it will be appropriate to still use a paired T-test

```

```{r}
# pH
group_by(pH_long, study_week) %>% #set groups to what we are comparing
summarise(sample_size = n()) #use summarise() and n() to get counts of ppl in each group

pH_wk1 <- pH_long %>%
  filter(study_week == "week1")
shapiro.test(pH_wk1$ph_mean) 
ggplot(pH_wk1, aes(x = ph_mean)) + geom_histogram()

pH_wk3 <- pH_long %>%
  filter(study_week == "week3")
shapiro.test(pH_wk3$ph_mean) 
ggplot(pH_wk3, aes(x = ph_mean)) + geom_histogram()

var.test(x = pH_wk1$ph_mean, 
         y = pH_wk3$ph_mean, 
         alternative = "two.sided")
#Both week 1 and week 3 for pH had similar sample sizes of well over 30. The week 1 data appears mostly normal, but slightly skewed to the left and passes the Shapiro-Wilk normality test with a p value of 0.09 which is greater than 0.05. The week 3 data is somewhat normal, but bimodal, and did not pass the Shapiro-Wilk normality test with a p value of 0.0036. Their variances were close to equal, with a p-value of 0.04, but ultimately fail that parameter. I think that it would still be reasonable to run a paired t-test, but I am going to run a non-parametric too just to be safe.  
```

# Homework 5.4
Conduct the appropriate statistical tests to determine if patterns observed in plots are significant. Write your interpretations of the results as a comment after the statistical tests.
```{r}
# breath methane
ch4_df <- inner_join(x = ch4_wk1, y = ch4_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(ch4_mean_wk1 = ch4_mean.x,
         ch4_mean_wk3 = ch4_mean.y) %>%
  select(-starts_with("study_week"))

wilcox.test(x = ch4_df$ch4_mean_wk1, 
            y = ch4_df$ch4_mean_wk3, 
            alternative = "less", paired = TRUE)
t.test(x = ch4_df$ch4_mean_wk1, 
       y = ch4_df$ch4_mean_wk3,
       alternative = "less", paired = TRUE, var.equal = FALSE)

#The p-values for both the non-parametric test and the paired t-test were low (p= 0.0024 and 0.008535 respectively), so we reject the null hypothesis that the concentration of methane did not change from week 1 to week 3 and support the alternative hypothesis that the concentration of methane increased from week 1 to week 3.
```

```{r}
# acetate
ace_df <- inner_join(x = ace_wk1, y = ace_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(acetate_mean_wk1 = acetate_mean.x,
         acetate_mean_wk3 = acetate_mean.y) %>%
  select(-starts_with("study_week"))


t.test(x = ace_df$acetate_mean_wk1, y = ace_df$acetate_mean_wk3, 
       alternative = "less", paired = TRUE) 

#The p-value for the t-test is very low (p=0.003459) which means that we reject the null hypothesis that the concentration of acetate did not change from week 1 to week 3 and supports the alternative hypothesis that the concentration of acetate increased from week 1 to week 3.
```

```{r}
# propionate
pro_df <- inner_join(x = pro_wk1, y = pro_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(propionate_mean_wk1 = propionate_mean.x,
         propionate_mean_wk3 = propionate_mean.y) %>%
  select(-starts_with("study_week"))

# code to run paired t-test 
t.test(x = pro_df$propionate_mean_wk1, y = pro_df$propionate_mean_wk3, 
       alternative = "less", paired = TRUE) 

#The p-value for the t-test is relatively high (p=0.3541) which means that we fail to reject the null hypothesis that the concentration of acetate did not change from week 1 to week 3.
```

```{r}
# pH
pH_df <- inner_join(x = pH_wk1, y = pH_wk3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed", "quantity_compliant")) %>%
  rename(ph_mean_wk1 = ph_mean.x,
         ph_mean_wk3 = ph_mean.y) %>%
  select(-starts_with("study_week"))

wilcox.test(x = pH_df$ph_mean_wk1, 
            y = pH_df$ph_mean_wk3, 
            alternative = "greater", paired = TRUE)
t.test(x = pH_df$ph_mean_wk1, 
       y = pH_df$ph_mean_wk3,
       alternative = "greater", paired = TRUE, var.equal = FALSE)
#The p-values for both the non-parametric test and the paired t-test were low (p= 0.02563 and 0.01013 respectively), so we reject the null hypothesis that the pH did not change from week 1 to week 3 and support the alternative hypothesis that the pH decreased from week 1 to week 3.
```


# ANOVA
As you know, we have more than two starch groups we would like to analyze. Instead of doing multiple pairwise comparisons with t-tests (which isn't the best approach because of reasons ...) we use an ANOVA, which compares all categorical groups to one another. 

To illustrate when an ANOVA would be useful, use the `scfa_wide` data frame imported above, plot the delta butyrate vs. supplement type. 
```{r}
scfa_wide %>%
  ggplot(aes(x = supplement_consumed,
             y = delta_butyrate,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Butyrate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Check assumptions

As with the t-tests, there are assumptions to check before running an ANOVA. 
```{r}
# check sample sizes
scfa_wide %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
```

Divide the labor with your neighbors and run the normality check on each of the 12 supplements. The first few have been completed as an example.
```{r}
# check normality of each group 
s1 <- scfa_wide %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$delta_butyrate) #p-value = 0.6886

s2 <- scfa_wide %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$delta_butyrate) #p-value = 0.7785

s3 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$delta_butyrate) #p-value = 0.4633

s4 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$delta_butyrate) #p-value = 0.7113

s5 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$delta_butyrate) #p-value = 0.6287

s6 <- scfa_wide %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$delta_butyrate) #p-value = 0.492

s7 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS")
shapiro.test(s7$delta_butyrate) #p-value = 0.2537

s8 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT")
shapiro.test(s8$delta_butyrate) #p-value = 0.6976

s9 <- scfa_wide %>%
  filter(supplement_consumed == "none")
shapiro.test(s9$delta_butyrate) #p-value = 0.009887

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s10$delta_butyrate) #p-value = 0.2783

s11 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS")
shapiro.test(s11$delta_butyrate) #p-value = 0.04406

s12 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize")
shapiro.test(s12$delta_butyrate) #p-value = 0.06542
```

```{r}
# check variances 
bartlett.test(delta_butyrate ~ supplement_consumed, data = scfa_wide)
# enter arguments with the following formula: continuous ~ categorical 
```

Now that we know our assumptions are reasonably met, we can run the test: 
```{r}
# run anova
aov_results <- aov(delta_butyrate ~ supplement_consumed, data = scfa_wide)
summary(aov_results)
```


# Homework 5.5
Repeat the processing of checking assumptions to conduct ANOVA on delta acetate and propionate. Create a plot for delta acetate and propionate. Save plots to folder called `figures`. Decide if you should proceed with conducting the ANOVA. Write your interpretations of the results as a comment after the statistical tests.
```{r}
# acetate
# sample sizes
scfa_wide %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())

# normality
s1 <- scfa_wide %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$delta_acetate) #p-value = 0.674

s2 <- scfa_wide %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$delta_acetate) #p-value = 0.05369

s3 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$delta_acetate) #p-value = 0.8305

s4 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$delta_acetate) #p-value = 0.001526

s5 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$delta_acetate) #p-value = 0.3026

s6 <- scfa_wide %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$delta_acetate) #p-value = 0.9105

s7 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS")
shapiro.test(s7$delta_acetate) #p-value = 0.02496

s8 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT")
shapiro.test(s8$delta_acetate) #p-value = 0.137

s9 <- scfa_wide %>%
  filter(supplement_consumed == "none")
shapiro.test(s9$delta_acetate) #p-value = 0.3468

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s10$delta_acetate) #p-value = 0.7713

s11 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS")
shapiro.test(s11$delta_acetate) #p-value = 0.9725

s12 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize")
shapiro.test(s12$delta_acetate) #p-value = 0.6092

# variances 
bartlett.test(delta_acetate ~ supplement_consumed, data = scfa_wide) #p-value = 1.167*10^-5

#plot
plot_anova_ace <- scfa_wide %>%
  ggplot(aes(x = supplement_consumed,
             y = delta_acetate,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Acetate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot_anova_ace

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_anova_ace.pdf",
          plot = plot_anova_ace,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# I belive that the assumptions are reasonably met, so it makes sense to move forward with the ANOVA test

# anova
aov_results <- aov(delta_acetate ~ supplement_consumed, data = scfa_wide)
summary(aov_results)

# Based on the ANOVA p value of 0.137, we fail to reject the null hypothesis that there is no statistically significant difference in the change in acetate concentration between different supplement types

```

```{r}
# propionate
# sample sizes
scfa_wide %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())

# normality
s1 <- scfa_wide %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$delta_propionate) #p-value = 6.507*10^-5

s2 <- scfa_wide %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$delta_propionate) #p-value = sample size not large enough

s3 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$delta_propionate) #p-value = 0.2266

s4 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$delta_propionate) #p-value = 0.2669

s5 <- scfa_wide %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$delta_propionate) #p-value = 0.4316

s6 <- scfa_wide %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$delta_propionate) #p-value = 0.8973

s7 <- scfa_wide %>%
  filter(supplement_consumed == "BRMPS")
shapiro.test(s7$delta_propionate) #p-value = 6.47*10^-6

s8 <- scfa_wide %>%
  filter(supplement_consumed == "LOODAT")
shapiro.test(s8$delta_propionate) #p-value = 0.8389

s9 <- scfa_wide %>%
  filter(supplement_consumed == "none")
shapiro.test(s9$delta_propionate) #p-value = 0.1025

s10 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium")
shapiro.test(s10$delta_propionate) #p-value = 0.1115

s11 <- scfa_wide %>%
  filter(supplement_consumed == "Psyllium+BRMPS")
shapiro.test(s11$delta_propionate) #p-value = 0.6636

s12 <- scfa_wide %>%
  filter(supplement_consumed == "transition_HiMaize")
shapiro.test(s12$delta_propionate) #p-value = 0.4711

# check variances 
bartlett.test(delta_propionate ~ supplement_consumed, data = scfa_wide)

#plot
plot_anova_pro <- scfa_wide %>%
  ggplot(aes(x = supplement_consumed,
             y = delta_propionate,
             color = supplement_consumed)) +
  geom_hline(yintercept = 0,
             linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Propionate mmol/kg (wk3 - wk1)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plot_anova_pro

save_plot(filename = "/Users/Owner/Documents/UMich_Bio201_F19/Lab5/figures/plot_anova_pro.pdf",
          plot = plot_anova_pro,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# even though some of the supplements have very low shapiro test values and the bartlett test for variances had a very low p value, I believe that the assumptions for an ANOVA test, though not ideal, are somewhat reasonably met, so I will run the test

# anova
aov_results <- aov(delta_propionate ~ supplement_consumed, data = scfa_wide)
summary(aov_results)

# Based on the ANOVA p value of 0.197, we fail to reject the null hypothesis that there is no statistically significant difference in the change in propionate concentration between different supplement types
```

