---
title: "Lab 6"
author: "Julia Van Goor"
date: "10/21/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Owner/Documents/UMich_Bio201_F19/Lab6/")
```

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
library(agricolae)
set.seed(7)
```

# Import data

### Import + format SCFA data

Import the SCFA data frame from last week. Update the columns names to match the course's style, set the order of the semesters by converting from character to factor. 

```{r include=FALSE}
scfa_long <- read_delim("C:/Users/Owner/Documents/UMich_Bio201_F19/Lab5/curated_data/SCFA_wkly_long.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower) %>%
  select(-ends_with("median")) %>% 
  mutate(totalscfa_mean = acetate_mean + butyrate_mean + propionate_mean,
         semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE)) %>%
  filter(quantity_compliant == "yes")
```

### Import tax data 

Import the taxonomic table `person_week_mean_by_cat_w_phylum.csv`. Notice the extension is csv, the delimitor in this file is a comma instead of a tab (which is different from what we have worked with previously). 
```{r}
# taxanomic data
taxa_category <- read_delim("raw_data/person_week_mean_by_cat_w_phylum.csv",
                            delim = ",", escape_double = FALSE, 
                            trim_ws = TRUE, na=c("NA"),
                            col_types = list())
View(taxa_category)
```

Look at the data frame, notice the week and semester formats are different then the rest of our data files. Today we are going to do some correlations between taxon abundance and short chain fatty acid (SCFA) concentrations. If we want to join these two data sets we will need make sure the format of these variables are the same. 

### Format tax data
```{r}
# create duplicate of data frame
taxa_cat_curated <- taxa_category %>%
  rename_all(tolower) %>%
  rename(participant_id = person,
         study_week = week,
         relative_abundance = fract) %>%
  select(participant_id, study_week, semester, tax_id, relative_abundance, phylum, tax_name, category)

# update week format 
taxa_cat_curated[["study_week"]] <- recode(.x=taxa_cat_curated[["study_week"]], 
                                  "week_1"="week1")
taxa_cat_curated[["study_week"]] <- recode(.x=taxa_cat_curated[["study_week"]], 
                                  "week_3"="week3")
# update semester format
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2015w"="Winter2015")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2015f"="Fall2015")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2016w"="Winter2016")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2016f"="Fall2016")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2017w"="Winter2017")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2017f"="Fall2017")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2018w"="Winter2018")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2018f"="Fall2018")
taxa_cat_curated[["semester"]] <- recode(.x=taxa_cat_curated[["semester"]], 
                                  "2019w"="Winter2019")

# Save a copy of the curated data frame
write_delim(taxa_cat_curated, path="curated_data/taxa_cat_curated.csv",
            delim = ",", quote = FALSE, col_names = TRUE)

View(taxa_cat_curated)
```

The cleaned-up `taxa_cat_curated` contains a subset of organisms found in the gut, specifically organisms which consume or produce metabolites of interest. The `tax_id` column is a number assigned to a sequence during the sequence processing pipeline. You won't need to use these right now. The `relative_abundance` column contains the frequencey of the organisms in each participant. The frequency is the weekly mean of (# sequences for each tax_id) / (total sequences per sample) for each participant. The `tax_name` is the most likely identity of each sequence. Information regarding microbial metabolism is in the `category` column:

* RS_degraders = Resistant starch degrading specialists (Bifidobacterium + Ruminococcus bromii)
* butyrogen_focused = A focused list of the most common / canonical butyrogens
* butyrogen_flint = A broader list of butyrogenic organisms (information comes from Harry Flint's research)
* hydrogen_consumers = Hydrogen consumers
* hydrogen_gen = Hydrogen generators
* propionate_gen = Propionate generators (essentially Bacteroides)

### Join SCFA + tax data

In the latter half of this lab we are going to test correlations. To test correlations between taxa and their metabolites, we need to join the data frames containing these data. Join the `tax_cat_curated` and `scfa_long` data frames. Remember to check the number of rows and participants in each data frame to help you select the best type of join. Discuss your decision with your neighbor, did you choose the same type of join? 

```{r}
#check number of rows
nrow(taxa_cat_curated) 
nrow(scfa_long) 

#check number of participants 
n_distinct(taxa_cat_curated$participant_id) 
n_distinct(scfa_long$participant_id) 

#identify shared columns = ?
colnames(taxa_cat_curated) 
colnames(scfa_long)
```

```{r}
tax_scfa <- inner_join(scfa_long, taxa_cat_curated,
                 by = c("participant_id", "study_week", "semester"))
```

# Intro

Last week we discussed the factors influencing data structure (sample size, varaince), how to check assumptions for statistical tests, how to determine if the differences between two means (t-tests, Mann-Whitney-U) or more (ANOVA) are likely real, and what to do if your datset does not meet the assumptions when comparing two means (non-parametric). 

In the first portion of the lab we will review the assumption checking process for comparison of more than 2 means, use the non-pararmetric counterpart to ANOVA (Kruskall-Wallace), and learn what to do if your ANOVA (or Kruskall-Wallace) test has a significant result. This will complete our section on categorical/discrete vs. continuous variables. The second half of the lab will discuss how to plot and analyze continuous vs. countinuous variables.

### Checking assumptions

Remember the assumptions for an ANOVA are similar to a t-test (large-ish sample size, normal distribution, equal variances). 

```{r eval=FALSE, include=FALSE}
# sample size of each group
group_by() %>% summarize(sample_size = n())

# check normality of each group
shaprio.test()
histogram()

# check variances
bartlett.test()
```

Last week each ANOVA you conducted should have resulted in a non-significant result (approx p >= 0.15). Run the code for the plot and ANOVA below. 

```{r}
# know assumptions are already met
scfa_long %>%
  filter(study_week == "week1") %>%
  ggplot(aes(x = semester,
             y = butyrate_mean, 
             color = semester)) +
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) +
  ylab("Week 1 butyrate (mmol/kg)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# ANOVA 
aov_wk1 <- aov(butyrate_mean ~ semester, scfa_long)
summary(aov_wk1)
```

Here we have a small p-value, this means at least one of our groups is significantly different from the others. To identify which group is different we have to run a pos-hoc test (an additional test). We will use Tukey's Honestly Significant Difference (HSD).

```{r}
# conduct Tukey's HSD test 
tuk_wk1 <- tidy(TukeyHSD(aov_wk1)) %>%
  mutate(adj.p.value = round(adj.p.value, digits = 2)) %>%
  filter(adj.p.value <= 0.10) #filter Tukey output for significant results 

# save data frame of results to file 
write_delim(tuk_wk1, path = "curated_data/tuk_wk1.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

### Kruskall-Wallace

If you have a data set that is clearly non-normal (such as the breath methane data), and want to ask a similar question as we asked with ANOVAs last week (e.g., Does the breath methane differ during week 3 for individuals consuming different supplements?), we need to use a non-parametric test. The non-parametric companion to an ANOVA is the Kruskall-Wallace test. 

Last week you should have checked assumptions on breath methane and determined that it was non-normal, so an ANOVA would not be an appropriate test to use to determine if the week 3 breath methane concentrations differed between supplements consumed.

```{r}
# import methane data 
breath_wkly <- read_delim("C:/Users/Owner/Documents/UMich_Bio201_F19/Lab5/curated_data/breath_wkly.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
# format methane data
  rename_all(tolower) %>%
  filter(study_week == "week3",
         quantity_compliant == "yes") %>% #remove non-compliant individuals
  mutate(semester = factor(semester,
                           levels=c("Winter2015", "Fall2015", 
                                    "Winter2016", "Fall2016", 
                                    "Winter2017", "Fall2017", 
                                    "Winter2018", "Fall2018", 
                                    "Winter2019"), ordered = TRUE))
# plot
breath_wkly %>%
  ggplot(aes(x = supplement_consumed, 
             y = ch4_mean, 
             color = supplement_consumed)) + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Breath methane (ppm)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# run the test
kruskal.test(supplement_consumed ~ ch4_mean, breath_wkly)
```

The interpretation for Kruskall-Wallace is similar to an ANOVA; small p-value indicates at least one group is different from the others. 

### Homework 6.1

* Write hypotheses to answer the question; do hydrogen breath levels differ between participants consuming fiber supplements? Create a plot of hydrogen levels:
```{r}
# hypothesis: 
# Null hypothesis: there is no statistically significant difference in hydrogen breath levels between participants consuming different fiber supplements
#Alternate hypothesis: there is a statistically significant difference in hydrogen breath levels between participants consuming different fiber supplements
# plot 
breath_wkly %>% 
  ggplot(aes(x = supplement_consumed, 
             y = h2_mean, 
             color = supplement_consumed)) + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Breath hydrogen (ppm)") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

* Which assumptions need checked for the statistical test(s)? Check the assumptions in the space below:
```{r}
# sample size of each group
breath_wkly %>% 
  group_by(supplement_consumed) %>% summarize(sample_size = n())

# check normality of each group
s1 <- breath_wkly %>% 
  filter(supplement_consumed == "Accessible")
  shapiro.test(s1$h2_mean) #p-value = 0.0937
  ggplot(s1, aes(x=h2_mean)) +
  geom_histogram() 
  
s2 <- breath_wkly %>% 
  filter(supplement_consumed == "Arabino")
  shapiro.test(s2$h2_mean) #p-value = 0.1695
  ggplot(s2, aes(x=h2_mean)) +
  geom_histogram()
  
s3 <- breath_wkly %>% 
  filter(supplement_consumed == "Banana")
  shapiro.test(s3$h2_mean) #p-value = 0.04114
  ggplot(s3, aes(x=h2_mean)) +
  geom_histogram()
  
s4 <- breath_wkly %>% 
  filter(supplement_consumed == "BRMPS")
  shapiro.test(s4$h2_mean) #p-value = 2.823e-12
  ggplot(s4, aes(x=h2_mean)) +
  geom_histogram()
  
s5 <- breath_wkly %>% 
  filter(supplement_consumed == "BRMPS+Accessible")
  shapiro.test(s5$h2_mean) #p-value = 0.05314
  ggplot(s5, aes(x=h2_mean)) +
  geom_histogram()
  
s6 <- breath_wkly %>% 
  filter(supplement_consumed == "HiMaize")
  shapiro.test(s6$h2_mean) #p-value = 9.956e-06
  ggplot(s6, aes(x=h2_mean)) +
  geom_histogram()
  
s7 <- breath_wkly %>% 
  filter(supplement_consumed == "HiMaize+BRMPS")
  shapiro.test(s7$h2_mean) #p-value = 0.001864
  ggplot(s7, aes(x=h2_mean)) +
  geom_histogram()
  
s8 <- breath_wkly %>% 
  filter(supplement_consumed == "Inulin")
  shapiro.test(s8$h2_mean) #p-value = 0.0002929
  ggplot(s8, aes(x=h2_mean)) +
  geom_histogram()
  
s9 <- breath_wkly %>% 
  filter(supplement_consumed == "LOODAT")
  shapiro.test(s9$h2_mean) #p-value = 7.089e-05
  ggplot(s9, aes(x=h2_mean)) +
  geom_histogram()
  
s10 <- breath_wkly %>% 
  filter(supplement_consumed == "Psyllium")
  shapiro.test(s10$h2_mean) #p-value = 0.4093
  ggplot(s10, aes(x=h2_mean)) +
  geom_histogram()
  
s11 <- breath_wkly %>% 
  filter(supplement_consumed == "Psyllium+BRMPS")
  shapiro.test(s11$h2_mean) #p-value = 0.472
  ggplot(s11, aes(x=h2_mean)) +
  geom_histogram()
  
s12 <- breath_wkly %>% 
  filter(supplement_consumed == "transition_HiMaize")
  shapiro.test(s12$h2_mean) #p-value = 7.102e-07
  ggplot(s12, aes(x=h2_mean)) +
  geom_histogram()

# check variances
bartlett.test(h2_mean ~ supplement_consumed, data = breath_wkly) #p-value = 6.83e-08
```

* Which statistical test(s) would you use to answer the question? Conduct the appropriate statistical test(s):
```{r}
#the ANOVA test and the Kruskal-Wallace test would both answer the question by comparing hydrogen levels across groups, but since the data is clearly non-normal, the Kruskal-Wallace test is the most appropriate

kruskal.test(supplement_consumed ~ h2_mean, breath_wkly) #p-value = 0.3001

```

* What is the answer to the question? Write your conclusions:
```{r}
# conclusion: The p-value for the Kruskal-Wallace test is high (p-value = 0.3001) so we fail to reject the null hyposthesis that there is no statistically significant difference in hydrogen breath levels between participants consuming different fiber supplements.
```

# Correlations

Correlations are used to determine if there is a relationship between two continuous variables. For the questions we are interested in, determining if there is a monotonic (or linear) relationship will provide insight. A monotonic relationship would occur when both continuous variables increase or decrease simultaneously, or when one variable increases the other decreases. 

To test correlations in this course, we are going to use the best practice approach of linear models (lm). However we do want to point out other correlation tests in case you encounter them elsewhere (Pearson which is parametric and Spearman Rank with is non-parametric). These tests have assumptions which are easy to violate, and the statistics provided (p and R values) do not have a straightfoward interpretation.

### Plot 2 continuous variables

To plot two continuous variables we will use a combination of geom_point() and geom_smooth(). 

```{r}
tax_scfa %>%
  filter(category == "butyrogen_focused",
         study_week == "week3",
         supplement_consumed == "HiMaize") %>%
  ggplot(aes(x = butyrate_mean,
             y = relative_abundance)) + 
  geom_point() + #puts data points to match x and y coordinates
  geom_smooth(method = "lm", #used to create a linear best fit line
              se = FALSE) + #hides confidence interval around line 
  xlab("Week 3 mean butyrate (mmol/kg)") + 
  ylab("Relative abundance") 
```

You plot two continuous variables and they appear to be correlated, you can determine if they are correlated by building a linear model. 

### Linear models

The linear model function is lm(). The linear models we will use have similar arguments as running an ANOVA. You must have the two variables (x and y) and a data frame in which each variable is a column. To access the results of the model, assign it to a new object. In the example below we will determine if there is a correlation between butyrate concentration and the abundance of some butyrate producing organisms. 

```{r}
m1 <- tax_scfa %>%
  filter(category == "butyrogen_focused", #subset dataframe to butyrate producers
         study_week == "week3", #subset for week3
         supplement_consumed == "HiMaize") %>% #subset for HiMaize consumers 
  lm(butyrate_mean ~ relative_abundance, data = .) #test relationship
summary(m1) #view results 
```

Look at the p-value in the same line as the F-statistic, this tells us if there is a significant result. This p-value is high (p-value = 0.2959), so we do not have significance. However if it was significant, we would then look at the adjusted R-squared value which indicates the quality of linear fit (higher R-squared is better >=0.8).

### Homework 6.2

Use the `tax_scfa` data frame, calculate the ratio of each SCFA in a new column (scfa/total). Check that your ratios sum to 1. 

```{r}
scfa_ratio <- tax_scfa %>%
  mutate(acetate_ratio = acetate_mean/totalscfa_mean) %>% 
  mutate(butyrate_ratio = butyrate_mean/totalscfa_mean) %>% 
  mutate(propionate_ratio = propionate_mean/totalscfa_mean) %>%
  mutate(ratio_check = acetate_ratio+butyrate_ratio+propionate_ratio)
```

Calculate the ratios of Firmicutes to Bacteroidetes (Firmicutes abund / Bacteroidetes abundance).

```{r}
# filter for Firmicutes, calculate sum per person per week (firmicutes_abund)
Fr <- scfa_ratio %>% 
  filter(phylum == "Firmicutes") %>% 
  group_by(participant_id, study_week) %>% 
  summarise(firmicutes_abund = sum(relative_abundance))

# filter for Bacteroidetes, calculate sum per person per week (bacteroidetes_abund)
Br <- scfa_ratio %>% 
  filter(phylum == "Bacteroidetes <phylum>") %>% 
  group_by(participant_id, study_week) %>% 
  summarise(bacteroidetes_abund = sum(relative_abundance))

# join subsetted data frames
tax_ratio <- inner_join(Fr, Br) %>%
  mutate(f_b_ratio = (firmicutes_abund / bacteroidetes_abund)) %>% #calculate ratios
  inner_join(scfa_ratio) #join with scfa_ratio data frame
```

### Homework 6.3
Determine if there is a relationship between the butyrate concentration and Firmicutes abundance. Conduct the statistical test and generate a plot. Save plot to folder within lab 6 called 'figures'.
```{r}
# plot
plot_firmicutes_butyrate <- tax_ratio %>%
  ggplot(aes(x = butyrate_mean,
             y = firmicutes_abund)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Mean butyrate (mmol/kg)") + 
  ylab("Firmicutes abundance") +
  labs(title="Butyrate conc. and Firmicutes abundance")

save_plot(filename = "figures/plot_firmicutes_butyrate.pdf",
          plot = plot_firmicutes_butyrate,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_ratio %>%
  lm(butyrate_mean ~ firmicutes_abund, data = .)

summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.003863
```

Do you draw the same conclusions if you use the butyrate ratio and Firmicutes to Bacteroidetes ratio (f_b_ratio)?
```{r}
# plot
plot_butratio_fbratio <- tax_ratio %>%
  ggplot(aes(x = butyrate_ratio,
             y = f_b_ratio)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Butyrate ratio") + 
  ylab("Firmicutes to Bacteroidetes ratio") +
  labs(title="Butyrate ratio to Firmicutes to Bacteroidetes ratio")

save_plot(filename = "figures/plot_butratio_fbratio.pdf",
          plot = plot_butratio_fbratio,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63b <- tax_ratio %>%
  lm(butyrate_ratio ~ f_b_ratio, data = .)

summary(m63b) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.02854
#conclusions: yes ultimately this p value and the adjusted R-squared value
```

### Homework 6.4
Determine if there is a relationship between the propionate concentration and Bacteroidetes abundance. Conduct the statistical test and generate a plot. Save the plot.
```{r}
# plot
plot_bacteroidetes_propionate <- tax_ratio %>%
  ggplot(aes(x = propionate_mean,
             y = bacteroidetes_abund)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Mean propionate (mmol/kg)") + 
  ylab("Bacteroidetes abundance") +
  labs(title="Propionate conc. and Bacteroidetes abundance")

save_plot(filename = "figures/plot_bacteroidetes_propionate.pdf",
          plot = plot_bacteroidetes_propionate,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_ratio %>%
  lm(propionate_mean ~ bacteroidetes_abund, data = .)

summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.04197

```

Do you draw the same conclusions if you use the propionate ratio and Firmicutes to Bacteroidetes ratio?
```{r}
# plot
plot_proratio_fbratio <- tax_ratio %>%
  ggplot(aes(x = propionate_ratio,
             y = f_b_ratio)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Propionate ratio") + 
  ylab("Firmicutes to Bacteroidetes ratio") +
  labs(title="Propionate ratio to Firmicutes to Bacteroidetes ratio")

save_plot(filename = "figures/plot_proratio_fbratio.pdf",
          plot = plot_proratio_fbratio,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63b <- tax_ratio %>%
  lm(propionate_ratio ~ f_b_ratio, data = .)

summary(m63b) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.0213
#conclusions: I would not draw the same conclusions as the previous test becuase even though the statistical test reveals similar results, the graph shows a slightly negative correlation because the ratio is of firmicutes to bacteroidetes (bacteroidetes on the bottom) which is the opposite of the propionate ratio (propionate on top)

```

### Homework 6.5
Import the curated pH+BSS data you generated during Lab4. Join the pH data frame with the tax_ratio data frame.
```{r}
# import 
pH_BSS_wkly <- pH_BSS_wkly <- read_delim("C:/Users/Owner/Documents/UMich_Bio201_F19/Lab4/curated_data/pH_BSS_wkly.CSV", 
    "\t", escape_double = FALSE, trim_ws = TRUE)
View(pH_BSS_wkly)

# join 
tax_scfa_pH <- inner_join(tax_ratio, pH_BSS_wkly,
                          by = c("participant_id", "study_week", "supplement_consumed")) %>%
  na.omit(semester)
```

Determine if there is a relationship between total SCFAs and acetate. Conduct the statistical test and generate a plot. Save the plot to figures folder.
```{r}
# plot
plot_scfas_ace <- tax_scfa_pH %>%
  ggplot(aes(x = acetate_mean,
             y = totalscfa_mean)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Mean acetate (mmol/kg)") + 
  ylab("Mean Total SCFA") +
  labs(title= "Mean Total SCFA and Acetate conc.")

save_plot(filename = "figures/plot_scfas_ace.pdf",
          plot = plot_scfas_ace,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_scfa_pH %>%
  lm(acetate_mean ~ totalscfa_mean, data = .)

summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.8704

```

Do you draw the same conclusions if you use the total SCFAs and acetate ratio?
```{r}
# plot
plot_scfas_aceratio <- tax_scfa_pH %>%
  ggplot(aes(x = acetate_ratio,
             y = totalscfa_mean)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Acetate Ratio") + 
  ylab("Mean Total SCFA") +
  labs(title= "Mean Total SCFA and Acetate Ratio")

save_plot(filename = "figures/plot_scfas_aceratio.pdf",
          plot = plot_scfas_aceratio,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_scfa_pH %>%
  lm(acetate_ratio ~ totalscfa_mean, data = .)

summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.01878
#Conclusion: Using the acetate ratio results in a much smaller/weaker correlation coefficient than the previous test
```

### Homework 6.6
Determine if there is a relationship between total SCFAs and pH. Conduct the statistical test and generate a plot. Save the plot.
```{r}
# plot
plot_scfas_pH <- tax_scfa_pH %>%
  ggplot(aes(x = rounded_avg_pH,
             y = totalscfa_mean)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Average pH") + 
  ylab("Mean Total SCFA") +
  labs(title= "Mean Total SCFA and Average pH")

save_plot(filename = "figures/plot_scfas_pH.pdf",
          plot = plot_scfas_pH,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

# correlation test 
m63a <- tax_scfa_pH %>%
  lm(rounded_avg_pH ~ totalscfa_mean, data = .)

summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.09942
```

### Homework 6.7
Determine if there is a relationship between butyrate and pH. Conduct the statistical test and generate a plot. Save the plot.
```{r}
plot_butyrate_pH <- tax_scfa_pH %>% 
  ggplot(aes(x = butyrate_mean,
             y = rounded_avg_pH)) +
  geom_point() +
  geom_smooth(method = "lm",
              se = FALSE) +
  xlab("Mean Butyrate Conc. (mmol/kg)") +
  ylab("Average pH") +
  labs(title = "Mean Butyrate Conc. and Avg. pH")

save_plot(filename = "figures/plot_butyrate_pH.pdf",
          plot = plot_butyrate_pH,
          nrow = 1, ncol = 1, 
          base_aspect_ratio = 1.1)

m63a <- tax_scfa_pH %>% 
  lm(butyrate_mean ~ rounded_avg_pH, data = .)


summary(m63a) #view results #p-value = 2.2e-16 #adjusted R-squared = 0.05209

```

### Homework 6.8
Save each data frame generated today:

* taxa_cat_curated
* scfa_ratio
* tax_scfa
* tax_scfa_pH

```{r}

write_delim(taxa_cat_curated, 
            path="curated_data/taxa_cat_curated.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(scfa_ratio, 
            path="curated_data/scfa_ratio.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(tax_scfa, 
            path="curated_data/tax_scfa.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
write_delim(tax_scfa, 
            path="curated_data/tax_scfa_pH.txt",
            delim = "\t", quote = FALSE, col_names = TRUE)
```

-----
end